Interspecies comparisons of Mg/Ca ratios in limpet shells

### 2024-06-24 Manuscript.qmd
added elemental background

---
title: "Interspecies comparisons of Mg/Ca ratios in limpet shells"
author:
  - name: Niklas Hausmann
    affiliations:
      - name: Leibniz Zentrum für Archäologie
        address: Ludwig-Lindenschmit-Forum 1
        city: Mainz
        country: Germany
        postal-code: 55116
    orcid: 0000–0002–4381–8334
    email: niklas@palaeo.eu
    attributes:
      corresponding: true
  - name: Donna Surge
    affiliations:
      - name: University of North Carolina
        address: 104 South Road, 225 Geology Building
        city: Chapel Hill, NC
        country: US
        postal-code: 27599-3315
  - name: Francisco Zangrando
    affiliations:
      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
        address: Avenida Maipú 305
        city: Ushuaia
        country: Argentina
        postal-code: V9410BJA
  - name: Angelica Tivoli
    affiliations:
      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
        address: Avenida Maipú 305
        city: Ushuaia
        country: Argentina
        postal-code: V9410BJA
  - name: Ivan Briz-Godino
    affiliations:
      - name: 
        address: 
        city: 
        country: Spain
        postal-code: 
abstract: |
  This study provides a short reassessment of the use of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures. While *Patella vulgata* along the Spanish shoreline has since then repeatedly produced reliable correlations between sea surface temperature and Mg/Ca ratios, this is not the case for other patelloid species. *Patella vulgata* and *Nacella deaureata* have been studied using Mg/Ca with mixed or contrary results. In this study, we present elemental maps of various such species together with stable oxygen isotope values for some of the specimens. Our dataset also includes specimens that were previously unsuccessful in providing significant correlations between δ^18^O and Mg/Ca ratios. By reassessing these previous specimens and including a wider range of  modern and archaeological samples from three patelloid species (*P. vulgata*, *N. deaureata*, and *N. magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change and serve as a means to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e. annual temperature minima and maxima) for the targeted analysis using δ^18^O or clumped oxygen isotope analysis.
keywords: 
  - Sclerochronology
  - Limpets
  - Elemental Ratio
  - Mg/Ca
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Palaeo^3^
      formatting: preprint
      model: 3p
      cite-style: authoryear
output:
  units: cm
---

@@

Problematically, links between Mg/Ca ratios and SST changes in limpets have also been shown to be unreliable [@Graniero2015-zv] and even where they were successful, anomalous patterns in some specimens had to be filtered out, reducing the overall robustness of the results of those successful studies [@Ferguson2011-zl].

While recent research particularly of *Patella* sp. in the Mediterranean and Southwest Europe have provided promising results [@Hausmann2019-fi; @Garcia-Escarzaga2015-jc; @Garcia-Escarzaga2018-nf]. There remains unclarity for Atlantic limpet species, particularly since it was last shown here, that they are not reliable recorders of palaeotemperature [@Graniero2015-zv].

@@

Relying and adding onto published datasets, allowed us to simultaneously avoid costs for new high-resolution δ^18^O-data, use real-world examples, and also provide pilot-data for areas of existing research interest. Generally establishing the usefulness and reliabilty of Mg/Ca as SST proxy in limpets should help to provide a platform for future research and an important stepping stone to better understand elemental ratios in other marine mollusc shells.

@@

In total, we studied 16 archaeological specimens. Of those, seven come from Quoygrew on Westray dating to 900 – 1200 CE. These shells have previously been studied to characterise seasonal temperature change during the Medieval Climate Anomaly and the Little Ice Age [@Surge2012-ba]. Together they show a range of lifespans, with some comparatively long for their size (QG1-7246-1, 12 years within 16 mm of growth) making them particularly susceptible to time-averaging. That said, the majority are under 5 years old. These shells are an ideal comparative dataset from an archaeological assemblage to compare to the modern shell ORK_LT5 [@Graniero2017-io].

Nine additional shells have been selected for LIBS analysis only, to provide a comparative dataset to the modern *Nacella* sp. shells. These have not been analysed for δ^18^O-values, but rather as a test to see whether patterns visible in the modern Mg/Ca ratio distributions are also preserved and apparent in the archaeological shells.

>>>>

---
title: "Interspecies comparisons of Mg/Ca ratios in limpet shells"
author:
  - name: Niklas Hausmann
    affiliations:
      - name: Leibniz Zentrum für Archäologie
        address: Ludwig-Lindenschmit-Forum 1
        city: Mainz
        country: Germany
        postal-code: 55116
    orcid: 0000–0002–4381–8334
    email: niklas@palaeo.eu
    attributes:
      corresponding: true
  - name: Donna Surge
    affiliations:
      - name: University of North Carolina
        address: 104 South Road, 225 Geology Building
        city: Chapel Hill, NC
        country: US
        postal-code: 27599-3315
  - name: Francisco Zangrando
    affiliations:
      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
        address: Avenida Maipú 305
        city: Ushuaia
        country: Argentina
        postal-code: V9410BJA
  - name: Angelica Tivoli
    affiliations:
      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
        address: Avenida Maipú 305
        city: Ushuaia
        country: Argentina
        postal-code: V9410BJA
  - name: Ivan Briz-Godino
    affiliations:
      - name: 
        address: 
        city: 
        country: Spain
        postal-code: 
abstract: |
  This study provides a short reassessment of the use of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures. While *Patella vulgata* along the Spanish shoreline has since then repeatedly produced reliable correlations between sea surface temperature and Mg/Ca ratios, this is not the case for other patelloid species. *Patella vulgata* and *Nacella deaureata* have been studied using Mg/Ca with mixed or contrary results. In this study, we present elemental maps of various such species together with stable oxygen isotope values for some of the specimens. Our dataset also includes specimens that were previously unsuccessful in providing significant correlations between δ^18^O and Mg/Ca ratios. By reassessing these previous specimens and including a wider range of  modern and archaeological samples from three patelloid species (*P. vulgata*, *N. deaureata*, and *N. magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change and serve as a means to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e. annual temperature minima and maxima) for the targeted analysis using δ^18^O or clumped oxygen isotope analysis.
keywords: 
  - Sclerochronology
  - Limpets
  - Elemental Ratio
  - Mg/Ca
date: last-modified
bibliography: bibliography.bib
format: html

  # elsevier-pdf:
  #   keep-tex: true
  #   journal:
  #     name: Palaeo^3^
  #     formatting: preprint
  #     model: 3p
  #     cite-style: authoryear
output:
  units: cm
---

@@

Problematically, links between Mg/Ca ratios and SST changes in many marine mollusc shells, including limpets, have shown to be unreliable [@Surge2008-ri; @Wanamaker2008-zl; @Schone2010-yl; @Freitas2012-tx; @Graniero2015-zv; @Poulain2015-dg; @Vihtakari2017-wd]. This is particularly the case where there is little available additional information on metabolic processes, organic components of the shell matrix, intra-increment and intra-shell variability, and growth rates, which can independently and unpredictably affect Mg/Ca ratios and confound their interpretation as temperature proxy. Confusingly, multiple different temperature equations have been found for the same species (see [@Freitas2012-tx; @Vihtakari2017-wd] and references therein). Coeval specimens sharing one locality can also show differences in their relation to SST [@Hausmann2019-fi]. Where the use of Mg/Ca as a palaeotemperature proxy was successful, anomalous patterns in some specimens had still to be filtered out by hand, reducing the overall robustness of the results of those successful studies [@Ferguson2011-zl].

While recent research particularly of *Patella* sp. in the Mediterranean and Southwest Europe have provided promising results [@Hausmann2019-fi; @Garcia-Escarzaga2015-jc; @Garcia-Escarzaga2018-nf]. There remains a lack of clarity for Atlantic limpet species, particularly since it was last shown here, that they are not reliable recorders of palaeotemperature [@Graniero2015-zv].

@@

By relying and adding onto published datasets, we were able to simultaneously avoid costs for new high-resolution δ^18^O-data, to use real-world examples, and to also provide pilot-data for areas of existing research interest. Generally establishing the usefulness and reliability of Mg/Ca as SST proxy in limpets should help to provide a platform for future research and an important stepping stone to better understand elemental ratios in other marine mollusc shells.

@@

In total, we studied 16 archaeological specimens. Of those, seven *P. vulgata* come from Quoygrew on Westray dating to 900 – 1200 CE. These shells have previously been studied to characterise seasonal temperature change during the Medieval Climate Anomaly and the Little Ice Age [@Surge2012-ba]. Together they show a range of lifespans, with some comparatively long for their size (QG1-7246-1, 12 years within 16 mm of growth) making them particularly susceptible to time-averaging. That said, the majority are under 5 years old. These shells are an ideal comparative dataset from an archaeological assemblage to compare to the modern shell ORK_LT5 [@Graniero2017-io].

Nine additional shells of *N. magellanica* (n=6) and *N*. *deaureata* (n=3) have been selected for LIBS analysis only, to provide a comparative dataset to the modern *Nacella* sp. shells. These have not been analysed for δ^18^O-values, but rather as a test to see whether patterns visible in the modern Mg/Ca ratio distributions are also preserved and apparent in the archaeological shells. They derive from the site of Heshkaia 35

### 2024-06-24 Manuscript.tex
added elemental background

Problematically, links between Mg/Ca ratios and SST changes in limpets
have also been shown to be unreliable \citep{Graniero2015-zv} and even
where they were successful, anomalous patterns in some specimens had to
be filtered out, reducing the overall robustness of the results of those
successful studies \citep{Ferguson2011-zl}.

>>>>

Problematically, links between Mg/Ca ratios and SST changes in many
marine mollusc shells, including limpets, have shown to be unreliable
\citep{Surge2008-ri, Wanamaker2008-zl, Schone2010-yl, Freitas2012-tx, Graniero2015-zv, Poulain2015-dg, Vihtakari2017-wd}.
This is particularly the case where there is little available additional
information on metabolic processes, organic components of the shell
matrix, intra-increment and intra-shell variability, and growth rates,
which can independently and unpredictably affect Mg/Ca ratios and
confound their interpretation as temperature proxy. Confoundingly,
multiple different temperature equations have been found for the same
species (see {[}\citet{Freitas2012-tx}; @Vihtakari2017-wd{]} and
references therein). Coeval specimens sharing one locality can also show
differences in their relation to SST \citep{Hausmann2019-fi}. Where the
use of Mg/Ca as a palaeotemperature proxy was successful, anomalous
patterns in some specimens had still to be filtered out by hand,
reducing the overall robustness of the results of those successful
studies \citep{Ferguson2011-zl}.
